---
title: "Example Notebook"
format: html
---

This code demonstrates how to apply the method, using the July sea surface
temperatures (SST) covariates and Central US temperatures example from the paper. 
In this example, we regress a time series of mean 2m air temperatures over the Central US against the SST covariates, looking for potential associations.

In order to apply the smoothing method, we need to fit the MLEs of the spatial
model parameters to the SST data. This can be done by running `example_mle_fit_script.R`.
 
The functions are written using the `future` package to enable easy
parallelization. If parallelization is not an option due to memory or core
constraints use `plan(sequential)` to run the script on a single core. If
using `plan(multicore)` make sure to run the script from the terminal using
`Rscript`.

```{r}
library(data.table)

# Plotting libraries
library(maps)
library(patchwork) # stacking ggplots
library(ggplot2)
```

`plan(multicore)` can be used to speed up the process if there are available
cores and memory. See scripts/application/sst_mle_fit_parallel.R for details
note that plan(multicore) should not be run from within Rstudio, use Rscript
with a .R script file instead.

```{r}
source("code/functions/process_data.R")
source("code/functions/dist_cov.R")
source("code/functions/cov_smooth.R")
source("code/functions/blockwise_MLE.R")
source("code/functions/inference.R")
```
```{r}
data_path <- "data/"
save_path <- "data/"

sst_filepath <- paste0(data_path, "sst_detrend_anom_jul.csv")
mle_filepath <- paste0(data_path, "sst_jul_mle_fit.csv")
t2m_path <- paste0(data_path, "t2m_GP_mean_jul.csv")

# Load the 2m air temperatures. These are the response variable of interest.
t2m_GP <- fread(t2m_path)
t2m_lm <- lm(t2m ~ time, data = t2m_GP)
t2m_anom <- t2m_lm$residuals

# Load the detrended July SST anomalies. 
# Only gridboxes with fewer than 25% of values being identical are kept.
# This arises at locations that are frozen over for a portion of the
# year.
sst_anoms <- load_obs_data(path = sst_filepath,
                           max_const_prop = 0.25)

sst_coord_dt <- extract_coord_dt(sst_anoms)

# standardize the anomalies and turn them into a data matrix
sst_mat_std <- extract_data_matrix(sst_anoms, "sst_anoms", scale = TRUE)

# calculates signed Zonal (East-West) and Meridional (North-South) distances
# For use in calculating the weights from the spatial model. 
sst_zonal_sdist <- find_signed_zonal_dist(sst_coord_dt)
sst_merid_sdist <- find_signed_merid_dist(sst_coord_dt)

loc_list <- get_loc_list(sst_coord_dt)
```

Once the MLEs are fit we can load in the MLE values of the parameters, and
use those to generate the smoothing weights, smooth the data, perform the
standard regression tests, and use Benjamini-Hochberg to find rejections.

```{r}
sst_jul_mle <- fread(mle_filepath)
mle_mat <- data.matrix(sst_jul_mle[, 1:3])

smoothing_weights <- find_smooth_weights(loc_list,
                                         zonal_sdist = sst_zonal_sdist,
                                         merid_sdist = sst_merid_sdist,
                                         cov_func = matern_1.5_cov,
                                         mle_params = mle_mat)

sst_jul_mat_smooth <- sst_mat_std %*% smoothing_weights

smooth_lm_fits <- fast_fit_lm_tstat(t2m_anom, sst_jul_mat_smooth)

n <- nrow(sst_jul_mat_smooth)
smooth_pvals <- 2 * pt(abs(smooth_lm_fits), df = n - 2, lower.tail = FALSE)
smooth_reject <- find_BH_reject(smooth_pvals, alpha = 0.1)
```

For comparison, we can also run Benjamini-Hochberg with the un-smoothed
covariates, and a regression analysis with no correction for multiple testing.

```{r}
base_lm_fits <- fast_fit_lm_tstat(t2m_anom, sst_mat_std)

base_pvals <- 2 * pt(abs(base_lm_fits), df = n - 2, lower.tail = FALSE)
base_reject <- find_BH_reject(base_pvals, alpha = 0.1)

base_reject$method <- "base"
smooth_reject$method <- "smooth"

reject_df <- rbind(base_reject, smooth_reject)

names(reject_df)[3] <- "reject"

no_BH <- data.table(location = 1:8195,
                    p_val = base_pvals,
                    reject = base_pvals < 0.05,
                    method = "no BH")

reject_df <- rbind(reject_df, no_BH)

fwrite(reject_df, file = paste0(save_path, "t2m_sst_rejections_jul.csv"))
```

```{r}
land_map <- sf::st_as_sf(map(wrap = c(0, 360), plot = FALSE, fill = TRUE))

t2m_rejections <- fread(paste0(data_path, 't2m_sst_rejections_jul.csv'))

t2m_rejections$method <- factor(t2m_rejections$method,
                                levels = c('no BH', 'base', 'smooth'),
                                labels = c('no FDR control', 'base', 'smooth'))

t2m_rejections <- t2m_rejections[sst_coord_dt, on = "location"]

no_fdr_rejections <- t2m_rejections[method == 'no FDR control']
base_rejections <- t2m_rejections[method == 'base']
smooth_rejections <- t2m_rejections[method == 'smooth']

no_fdr_plot <- ggplot(data = land_map) +
  geom_sf(fill = "gray",
          color = "gray",
          inherit.aes = FALSE) +
  geom_point(data = no_fdr_rejections[reject == TRUE],
             size = 0.4,
             color = '#5A5A5A',
             aes(x = lon, y = lat)) +
  coord_sf(xlim = c(0, 360),
           ylim = c(-60, 60),
           expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 360, by = 50),
                     labels =  \(x) as.character(x)) +
  scale_y_continuous(breaks = seq(-60, 60, by = 20),
                     labels = \(x) as.character(x)) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_bw() +
  labs(title = 'a) no FDR control') + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        title = element_text(size = 8))

base_plot <- ggplot(data = land_map) +
  geom_sf(fill = "gray",
          color = "gray",
          inherit.aes = FALSE) +
  geom_point(data = base_rejections[reject == TRUE],
             size = 0.4,
             color = '#5A5A5A',
             aes(x = lon, y = lat)) +
  coord_sf(xlim = c(0, 360),
           ylim = c(-60, 60),
           expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 360, by = 50),
                     labels =  \(x) as.character(x)) +
  scale_y_continuous(breaks = seq(-60, 60, by = 20),
                     labels = \(x) as.character(x)) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_bw() +
  labs(title = 'b) base') + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        title = element_text(size = 8))

smooth_plot <- ggplot(data = land_map) +
  geom_sf(fill = "gray",
          color = "gray",
          inherit.aes = FALSE) +
  geom_point(data = smooth_rejections[reject == TRUE],
             size = 0.4,
             color = '#5A5A5A',
             aes(x = lon, y = lat)) +
  coord_sf(xlim = c(0, 360),
           ylim = c(-60, 60),
           expand = FALSE) +
  scale_x_continuous(breaks = seq(0, 360, by = 50),
                     labels =  \(x) as.character(x)) +
  scale_y_continuous(breaks = seq(-60, 60, by = 20),
                     labels = \(x) as.character(x)) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  labs(title='c) smooth') + 
  theme_bw() +
  theme(title = element_text(size = 8))

patchwork_stacked_plot <- (no_fdr_plot / base_plot / smooth_plot)
patchwork_stacked_plot
```